When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-12-05 code/Tahkv1.Rmd
commit

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
            pdftitle={When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America},
            pdfauthor={Peter M. Erickson},
            colorlinks=true,
            linkcolor=magenta,
            filecolor=Maroon,
            citecolor=black,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin = 1.15in]{geometry}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% there are a few ways to include latex options.
% 1: as a preamble file (like this one)
% 2: directly into the header-includes YAML variable (global)
% 3: output templates. For more info on templates,
%    see <https://bookdown.org/yihui/rmarkdown/rticles-templates.html>
%    and <https://bookdown.org/yihui/rmarkdown/rticles-usage.html>

% These packages control the fonts used in PDF output.
% The standard LaTeX engine (PDFLaTeX) has a limited selection of fonts.
% See <https://www.tug.org/FontCatalogue/>

\usepackage{libertine}                 % serif font
\usepackage{libertinust1math}          % matching math font
\usepackage[scaled = 0.95, varqu]{zi4} % teletype (monospace/code) font

% The XeLaTeX engine gives you more control to specify fonts
%   at the expense of having poorer typesetting principles
%   (but xelatex typesetting is getting better)

% this package enables decimal-aligned table columns
\usepackage{dcolumn}

% the default PDF template for R Markdown loads a lot of convenient packages
% so you don't actually have to mess with them all that much.
\usepackage{setspace}
\usepackage{multirow}
\usepackage{indentfirst}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[style=authoryear,]{biblatex}
\addbibresource{researchpaper-ps904.bib}

\title{When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America}
\author{Peter M. Erickson}
\date{December 03, 2020}

\begin{document}
\maketitle
\begin{abstract}
\emph{This chapter explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers since 1979. Drawing on an original data set almost 400 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold the concept of military professionalism as defined by historical military norms and standards. What explains this variation? The author argues that increases in political polarization and the prestige of the military account for the increase in the number of op-eds which challenge traditional conceptions and norms of military professionalism. The paper concludes by arguing that the concept of ``politicization of the military'' --- something that previous civil-military relations scholars have identified and described --- must be widened to account not only for efforts by politicians to use the military for political advantages, but also for the phenomenon of retired military officers engaging the public to achieve political ends. The author argues that such a development, if uncorrected, will cause significant damage to the military as a profession.}\\
\end{abstract}

@@

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In some cities and communities across the nation, peaceful protests were followed by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St.~John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House. In the days and weeks that followed, a civil-military relations crisis ensued \autocite{feaver_military_2020}. The Secretary of Defense, Mark Esper, publicly disagreed with the President's threat to invoke the Insurrection Act of 1807 to use active-duty forces to quell domestic riots \autocite{esper_secretary_2020}, and General Mark Milley, the Chairman of the Joint Chiefs of Staff, later told senior military officers graduating from the National Defense University that he ``should not have been there'' (accompanying the President on the walk to St.~John's Church), as his presence ``created a perception of the military involved in domestic politics'' \autocite{milley_official_2020}.

Furthermore, several retired military officers weighed in not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America.\footnote{These officers included several retired four star generals and admirals}. However, there was marked variation in the tone and tenor of the remarks made by these officers: although each expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution \autocites{brooks_dismay_2020}{dempsey_former_2020}, some officers went much further in their criticisms, leveling direct insults of the President, including his fitness to lead \autocite{goldberg_james_2020}, his administration's broader policies on race in America \autocite{allen_moment_2020}, and his general leadership abilities \autocite{mullen_i_2020}. Moreover, the comments made by these retired senior leaders renewed interest, at least for a while, in the topic of retired military officer speaking out on current events - a topic that civil-military relations scholars have debated \autocites{kagan_let_2006}{brooks_let_2020}, but one that has yet, in the author's opinion, to be studied empirically and systematically. In a domestic political age in which polarization is high, the topic of retired military officer speech is important for a variety of reasons. Not only does the speech of retired military officers carry the potential to impact public opinion, but the very act of retired senior leaders speaking out may also compel those who are still in the military to follow suit. More importantly, the topics that military leaders choose to address in the public square, and the tone of their remarks, is reflective of the conceptualizations these officers have regarding their role in society and ultimately, their purpose and professional identity as military officers. Indeed, in 2020, retired as well as active duty military officers grabbed the pen and argued in support of a variety of topics that clearly pushed the traditional boundaries of professional conduct by military officers by openly arguing for a variety of topics related to President Donald Trump, including including race in America \autocite{askew_anti-racist_2020} and even removing the President from the White House \autocite{nagl__2020}.

@@

This paper proceeds in three parts. First, the paper connects the concept of military ``role beliefs'' to retired military officer speech. Second, the paper explores and analyzes an original data set of nearly 390 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. This analysis ultimately demonstrates that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen both to argue about topics that fall outside of the realm of traditional national security topics, and to violate the principle of civilian control of the military when doing so. The analysis also shows that these trends have worsened as the American political climate has become increasingly polarized over the past several decades. Third, thechapter concludes by exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a highly polarized setting that will likely endure.

\singlespace

\hypertarget{how-the-military-views-itself}{%
\section{How the Military Views Itself}\label{how-the-military-views-itself}}

\hypertarget{military-role-beliefs}{%
\subsection{Military Role Beliefs}\label{military-role-beliefs}}

The concept of military ``role beliefs'' was introduced by J. Samuel Fitch in his book, \emph{The Armed Forces and Democracy in Latin America.} In the book, Fitch describes military ``role beliefs'' as the perceptions held by military officers regarding their normative role in politics \autocite[61]{fitch_armed_1998}. Though Fitch's focus is on militaries in Latin America, it is possible to use the concept of ``role beliefs'' to develop a normative starting point applicable to military officers in the United States.

Two foundations for developing a starting point for military ``role beliefs'' in the United States are the tandem norms of civilian control of the military, and the idea of the military acting as a non-partisan institution. Though such norms have been episodically challenged throughout American history in a variety of ways and during specific instances, these norms have nonetheless been both foundational and aspirational, particularly in the post World War Two period \autocite{weigley_american_2001}. The norm of civilian control of the military stems directly from the US Constitution, which expressly grants control of the military (as well as specific responsibilities associated with maintaining control) to the President, Congress, and the individual states \autocite{noauthor_us_1787-2}. Unlike the norm of civilian control, the norm of the military functioning as a non-partisan institution is not expressly spelled out in the Constitution, but rather the product of the American military's collective experiences and traditions in fighting and prevailing in the nation's wars and conflicts.

Further, a number of scholars have underscored the importance of these two norms, and the relationship between these norms and the military's ability to prevail during times of war. In his influential book, \emph{The Soldier and the State: The Theory and Politics of Civil-Military Relations}, Samuel Huntington argues in favor of maintaining separate operating spheres between the political leaders of a democracy and the military as an institution in order to ensure civilian control of the military by rendering the military ``politically sterile and neutral,'' or non-partisan \autocite[84]{huntington_soldier_1957}. Huntington goes on to argue that only by such an arrangement, which he called ``objective control of the military,'' will the military be capable of defeating significant security threats while also remaining apolitical \autocite[83-84]{huntington_soldier_1957}.

Having established the foundational norms of civilian control of the military and the military operating as an apolitical institution, the next step is to extrapolate how these norms might impact the public speech of military officers. This requires balancing these norms with the fact that military officers are nonetheless servants of the Republic who have accumulated knowledge and experiences over the course of sometimes lengthy careers (in the case of retired military officers, at least 20 and in some cases, upwards of 40 or even more years of experience), and thus, in many cases are qualified to weigh in on circumstances and situations that impact the nation. For instance, in the aftermath of the ``Revolt of the Generals'' in 2006, in which several retired military officers publicly criticized and called for the removal of Defense Secretary Donald Rumsfeld, Frederick Kagan argued in support of retired military officers having the freedom to weigh in policy debates, particularly those involving an active armed conflict, stating, ``there is no danger to the republic in a handful of retired generals speaking their minds. There is great danger in making vital decisions about an on-going armed struggle without hearing the views of all available experts'' \autocite{kagan_let_2006}. Kagan's remarks lend support to the notion that in democracies in particular, transparent and flowing information is a precious resource that helps leaders make decisions and, more importantly, the public to gain awareness of the state's actions {[}need cite{]}.

Nonetheless, there have been both formal and informal boundaries placed on various forms of political activities by military officers. Formally speaking, Department of Defense (DoD) Directive 1344.10 contains a ``standing list of political behavior do's and don'ts for members of the active duty military'' \autocite[1]{urben_wearing_2014}. This list includes authorizing members of the military to engage in activities such as voting and joining a partisan or non-partisan club and to attend this club's meetings when not in uniform, and prohibits members of the military from activities such as participating in partisan fundraising, or advocating for or against a partisan cause or figure on radio or television \autocite{department_of_defense_political_2008}. Informally, a number of recent Chairmen of the Joint Chiefs of Staff have likewise urged members of the military to refrain from engaging in overly political activities, particularly in election years, when such activity is more likely to negatively reflect on the military. For example, prior to the 2016 Presidential Election, then-Chairman of the Joint Chiefs of Staff General Joseph Dunford reminded the military that the basis upon which it is able to maintain the public's trust is it's ability to remain ``apolitical and neutral'' \autocite{dunford_upholding_2016}.

Based on these formal and informal prohibitions on the political activities conducted by military officers, it is possible to conclude that visible behaviors that are overtly partisan are to be avoided at all costs, and are inappropriate for military officers to engage in. It is also possible to extrapolate that the norm of civilian control of the military results in a corresponding prohibition on members of the military criticizing political leaders to the point that civilian control of the military is jeopardized or fails to be upheld. Where this line should be drawn in practice is not clear, however, because as Kagan has pointed out, criticisms can and often do point out blind spots in a plan or otherwise reveal shortcomings that may not otherwise have been identified. However, there is a point at which a criticism levied at a political leader by a military officer goes beyond criticism. Finally, it is possible to conclude that activities which build on the appropriate expertise and knowledge of the military officer are appropriate, provided that these activities do not violate the stipulations previously mentioned, i.e.~a non-partisan military, and civilian control of the military.

\hypertarget{opinion-editorial-publication-as-a-political-activity}{%
\subsection{Opinion-Editorial Publication as a Political Activity?}\label{opinion-editorial-publication-as-a-political-activity}}

We can now focus on the actual behavior of the military authoring op-ed pieces. Should these considered a form of political activity? I argue that it depends on what message the op-ed actually sends or is intended to send. An op-ed that strictly points out the pitfalls of the nation's combat strategy written by an experienced military officer may not be inappropriate, whereas an op-ed written by a retired military officer that argues in favor of a specific Presidential candidate is, at least by the standards put forth in this paper, entirely inappropriate. Therefore, at least two questions should be asked when examining an opinion piece authored by a retired military officer: first, does the piece violate either the norm of civilian control of the military or the military as an apolitical institution, and second, does the op-ed address an appropriate topic that a military officer is qualified to address?

Beyond these two questions, however, opinion-editorials are important because they serve as an informative barometer of how the military institution sees itself, and particularly, its role with respect to the American public. Members of the active military do not frequently write op-eds, but there have been notable exceptions over the years, perhaps none so notable as several written by then-Chairman of the Joint Chiefs of Staff General Colin Powell in the early 1990s. For instance, Powell presented the public with his view regarding the potential use of American military force in several developing ``hot spots'' across the world, including the Balkans, Somalia, and other places, during an October 8, 1992 piece entitled, ``Why Generals Get Nervous.'' Published in the \emph{New York Times}, Powell wrote in part, ``decisive means and results are always to be preferred, even if they are not always possible. So you bet I get nervous when so-called experts suggest that all we need is a little surgical bombing or a limited attack. When the desired result isn't obtained, a new set of experts then comes forward with talk of a little esclalation. History has not been kind to this approach'' \autocite{powell_opinion_1992}.

Powell clearly felt compelled to inform the public of how he, as the military's top officer, viewed the potential use of American military force abroad during the prevailing international context at the time. What is less known, however, is that Powell was also writing against a domestic backdrop in which there was an election coming up in less than a month that pitted the incumbent, George W. Bush, against Democratic challenger Bill Clinton. The two candidates differed with respect to foreign policy on a number of issues, and Bill Clinton was not highly respected by the military establishment, a fact that would impact the military's relationship with Clinton during his early years in office {[}need cite{]}. Whether it was proper for Powell to write such an op-ed truly depends: one could argue that the nation's top general was simply informing the public based on his years of individual experience serving in the military, in which case, from a democratic transparency standpoint, the op-ed well served the Republic; alternately, one could also argue that by writing during an election year, Powell's remarks brazenly staked the claims of the military ahead of a potential Presidential transition (which turned out to be the case), and in the process, limited the President's options for potentially using force abroad during his early days in office. The first explanation argues that from the standpoint of civil-military relations, there was no problem with Powell's opinion-editorial, while the second argues that the op-ed was nothing but problematic.

\hypertarget{retired-military-officers}{%
\subsection{Retired Military Officers}\label{retired-military-officers}}

One might argue that the logic employed thus far is seemingly plausible for members of the active military, but that the cautionary tone and reservations regarding certain types of political activity should not apply for members of the retired military community. And indeed, there are obvious distinctions between those who are on active duty and those who have retired from the service. These distinctions are made clear in fact in portions of DoD 1344.10 \autocite{department_of_defense_political_2008}. But the fact of the matter is that, especially for retired military officers who achieved general or flag officer rank, these retired officers sometimes carry outsized influence and impact when they engage in activities that approach on the political. During the Presidential Campaign of 2016, Retired Army General Martin Dempsey published an opinion-editorial piece in \emph{The Washington Post} which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the respective Republican and Democratic conventions. Dempsey wrote, ``As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors -- who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so'' \autocite{dempsey_military_2016}.

@@

\begin{quote}
\textbf{Yours is the profession of arms, the will to win, the sure knowledge that in war there is no substitute for victory, that if you lose, the Nation will be destroyed, that the very obsession of your public service must be Duty, Honor, Country. Others will debate the controversial issues, national and international, which divide men's minds. But serene, calm, aloof, you stand as the Nation's war guardians, as its lifeguards from the raging tides of international conflict, as its gladiators in the arena of battle. For a century and a half you have defended, guarded and protected its hallowed traditions of liberty and freedom, of right and justice. Let civilian voices argue the merits or demerits of our processes of government. Whether our strength is being sapped by deficit financing indulged in too long, by federal paternalism grown too mighty, by power groups grown too arrogant, by politics grown too corrupt, by crime grown too rampant, by morals grown too low, by taxes grown too high, by extremists grown too violent; whether our personal liberties are as firm and complete as they should be. These great national problems are not for your professional participation or military solution. Your guidepost stands out like a tenfold beacon in the night: Duty, Honor, Country} \autocite{macarthur_duty_1962}.
\end{quote}

>>>>

---
title: "When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This chapter explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers since 1979. Drawing on an original data set almost 400 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold the concept of military professionalism as defined by historical military norms and standards. What explains this variation? The author argues that increases in political polarization and the prestige of the military account for the increase in the number of op-eds which challenge traditional conceptions and norms of military professionalism. The paper concludes by arguing that the concept of "politicization of the military" --- something that previous civil-military relations scholars have identified and described --- must be widened to account not only for efforts by politicians to use the military for political advantages, but also for the phenomenon of retired military officers engaging the public to achieve political ends. The author argues that such a development, if uncorrected, will cause significant damage to the military as a profession._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

header-includes:
    - \usepackage{setspace}
    - \usepackage{multirow}
    - \usepackage{indentfirst}
    - \usepackage{caption}

output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: pdflatex
    number_sections: false
    toc: false
    fig_caption: true
    citation_package: biblatex
    includes: 
      in_header: 
        - latex-preamble.tex
  bookdown::htmldocument2: default
  bookdown::word_document2:
      toc: false
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("stargazer")
library("rstatix")
```
\newpage

# Introduction

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In some cities and communities across the nation, peaceful protests were followed by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House. In the days and weeks that followed, a civil-military relations crisis ensued  [@feaver_military_2020]. The Secretary of Defense, Mark Esper, publicly disagreed with the President's threat to invoke the Insurrection Act of 1807 to use active-duty forces to quell domestic riots [@esper_secretary_2020], and General Mark Milley, the Chairman of the Joint Chiefs of Staff, later told senior military officers graduating from the National Defense University that he "should not have been there" (accompanying the President on the walk to St. John's Church), as his presence "created a perception of the military involved in domestic politics" [@milley_official_2020].

Furthermore, several retired military officers weighed in not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America.^[These officers included several retired four star generals and admirals]. However, there was marked variation in the tone and tenor of the remarks made by these officers: although each expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticisms, leveling direct insults of the President, including his fitness to lead [@goldberg_james_2020], his administration's broader policies on race in America [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020]. Moreover, the comments made by these retired senior leaders renewed interest, at least for a while, in the topic of retired military officer speaking out on current events - a topic that civil-military relations scholars have debated [@kagan_let_2006; @brooks_let_2020], but one that has yet, in the author's opinion, to be studied empirically and systematically. In a domestic political age in which polarization is high, the topic of retired military officer speech is important for a variety of reasons. Not only does the speech of retired military officers carry the potential to impact public opinion, but the very act of retired senior leaders speaking out may also compel those who are still in the military to follow suit. More importantly, the topics that military leaders choose to address in the public square, and the tone of their remarks, is reflective of the conceptualizations these officers have regarding their role in society and ultimately, their purpose and professional identity as military officers. Indeed, in 2020, retired as well as active duty military officers grabbed the pen and argued in support of a variety of topics that clearly pushed the traditional boundaries of professional conduct by military officers by openly arguing for a variety of topics related to President Donald Trump, including including race in America [@askew_anti-racist_2020] and even removing the President from the White House [@nagl__2020].

@@

This paper proceeds in three parts. First, the paper connects the concept of military "role beliefs" to retired military officer speech. Second, the paper explores and analyzes an original data set of nearly 390 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. This analysis ultimately demonstrates that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen both to argue about topics that fall outside of the realm of traditional national security topics, and to violate the principle of civilian control of the military when doing so. The analysis also shows that these trends have worsened as the American political climate has become increasingly polarized over the past several decades. Third, thechapter concludes by exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a highly polarized setting that will likely endure.

\singlespace
# How the Military Views Itself
\doublespace
## Military Role Beliefs
The concept of military "role beliefs" was introduced by J. Samuel Fitch in his book, _The Armed Forces and Democracy in Latin America._ In the book, Fitch describes military "role beliefs" as the perceptions held by military officers regarding their normative role in politics [@fitch_armed_1998, 61]. Though Fitch's focus is on militaries in Latin America, it is possible to use the concept of "role beliefs" to develop a normative starting point applicable to military officers in the United States.

Two foundations for developing a starting point for military "role beliefs" in the United States are the tandem norms of civilian control of the military, and the idea of the military acting as a non-partisan institution. Though such norms have been episodically challenged throughout American history in a variety of ways and during specific instances, these norms have nonetheless been both foundational and aspirational, particularly in the post World War Two period [@weigley_american_2001]. The norm of civilian control of the military stems directly from the US Constitution, which expressly grants control of the military (as well as specific responsibilities associated with maintaining control) to the President, Congress, and the individual states [@noauthor_us_1787-2]. Unlike the norm of civilian control, the norm of the military functioning as a non-partisan institution is not expressly spelled out in the Constitution, but rather the product of the American military's collective experiences and traditions in fighting and prevailing in the nation's wars and conflicts.

Further, a number of scholars have underscored the importance of these two norms, and the relationship between these norms and the military's ability to prevail during times of war. In his influential book, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, Samuel Huntington argues in favor of maintaining separate operating spheres between the political leaders of a democracy and the military as an institution in order to ensure civilian control of the military by rendering the military "politically sterile and neutral," or non-partisan [@huntington_soldier_1957, 84]. Huntington goes on to argue that only by such an arrangement, which he called "objective control of the military," will the military be capable of defeating significant security threats while also remaining apolitical [@huntington_soldier_1957, 83-84].

Having established the foundational norms of civilian control of the military and the military operating as an apolitical institution, the next step is to extrapolate how these norms might impact the public speech of military officers. This requires balancing these norms with the fact that military officers are nonetheless servants of the Republic who have accumulated knowledge and experiences over the course of sometimes lengthy careers (in the case of retired military officers, at least 20 and in some cases, upwards of 40 or even more years of experience), and thus, in many cases are qualified to weigh in on circumstances and situations that impact the nation. For instance, in the aftermath of the "Revolt of the Generals" in 2006, in which several retired military officers publicly criticized and called for the removal of Defense Secretary Donald Rumsfeld, Frederick Kagan argued in support of retired military officers having the freedom to weigh in policy debates, particularly those involving an active armed conflict, stating, "there is no danger to the republic in a handful of retired generals speaking their minds. There is great danger in making vital decisions about an on-going armed struggle without hearing the views of all available experts" [@kagan_let_2006]. Kagan's remarks lend support to the notion that in democracies in particular, transparent and flowing information is a precious resource that helps leaders make decisions and, more importantly, the public to gain awareness of the state's actions [need cite].

Nonetheless, there have been both formal and informal boundaries placed on various forms of political activities by military officers. Formally speaking, Department of Defense (DoD) Directive 1344.10 contains a "standing list of political behavior do's and don'ts for members of the active duty military" [@urben_wearing_2014, 1]. This list includes authorizing members of the military to engage in activities such as voting and joining a partisan or non-partisan club and to attend this club's meetings when not in uniform, and prohibits members of the military from activities such as participating in partisan fundraising, or advocating for or against a partisan cause or figure on radio or television [@department_of_defense_political_2008]. Informally, a number of recent Chairmen of the Joint Chiefs of Staff have likewise urged members of the military to refrain from engaging in overly political activities, particularly in election years, when such activity is more likely to negatively reflect on the military. For example, prior to the 2016 Presidential Election, then-Chairman of the Joint Chiefs of Staff General Joseph Dunford reminded the military that the basis upon which it is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016].

Based on these formal and informal prohibitions on the political activities conducted by military officers, it is possible to conclude that visible behaviors that are overtly partisan are to be avoided at all costs, and are inappropriate for military officers to engage in. It is also possible to extrapolate that the norm of civilian control of the military results in a corresponding prohibition on members of the military criticizing political leaders to the point that civilian control of the military is jeopardized or fails to be upheld. Where this line should be drawn in practice is not clear, however, because as Kagan has pointed out, criticisms can and often do point out blind spots in a plan or otherwise reveal shortcomings that may not otherwise have been identified. However, there is a point at which a criticism levied at a political leader by a military officer goes beyond criticism. Finally, it is possible to conclude that activities which build on the appropriate expertise and knowledge of the military officer are appropriate, provided that these activities do not violate the stipulations previously mentioned, i.e. a non-partisan military, and civilian control of the military.

## Opinion-Editorial Publication as a Political Activity?
We can now focus on the actual behavior of the military authoring op-ed pieces. Should these considered a form of political activity? I argue that it depends on what message the op-ed actually sends or is intended to send. An op-ed that strictly points out the pitfalls of the nation's combat strategy written by an experienced military officer may not be inappropriate, whereas an op-ed written by a retired military officer that argues in favor of a specific Presidential candidate is, at least by the standards put forth in this paper, entirely inappropriate. Therefore, at least two questions should be asked when examining an opinion piece authored by a retired military officer: first, does the piece violate either the norm of civilian control of the military or the military as an apolitical institution, and second, does the op-ed address an appropriate topic that a military officer is qualified to address?

Beyond these two questions, however, opinion-editorials are important because they serve as an informative barometer of how the military institution sees itself, and particularly, its role with respect to the American public. Members of the active military do not frequently write op-eds, but there have been notable exceptions over the years, perhaps none so notable as several written by then-Chairman of the Joint Chiefs of Staff General Colin Powell in the early 1990s. For instance, Powell presented the public with his view regarding the potential use of American military force in several developing "hot spots" across the world, including the Balkans, Somalia, and other places, during an October 8, 1992 piece entitled, "Why Generals Get Nervous." Published in the _New York Times_, Powell wrote in part, "decisive means and results are always to be preferred, even if they are not always possible. So you bet I get nervous when so-called experts suggest that all we need is a little surgical bombing or a limited attack. When the desired result isn't obtained, a new set of experts then comes forward with talk of a little esclalation. History has not been kind to this approach" [@powell_opinion_1992].

Powell clearly felt compelled to inform the public of how he, as the military's top officer, viewed the potential use of American military force abroad during the prevailing international context at the time. What is less known, however, is that Powell was also writing against a domestic backdrop in which there was an election coming up in less than a month that pitted the incumbent, George W. Bush, against Democratic challenger Bill Clinton. The two candidates differed with respect to foreign policy on a number of issues, and Bill Clinton was not highly respected by the military establishment, a fact that would impact the military's relationship with Clinton during his early years in office [need cite]. Whether it was proper for Powell to write such an op-ed truly depends: one could argue that the nation's top general was simply informing the public based on his years of individual experience serving in the military, in which case, from a democratic transparency standpoint, the op-ed well served the Republic; alternately, one could also argue that by writing during an election year, Powell's remarks brazenly staked the claims of the military ahead of a potential Presidential transition (which turned out to be the case), and in the process, limited the President's options for potentially using force abroad during his early days in office. The first explanation argues that from the standpoint of civil-military relations, there was no problem with Powell's opinion-editorial, while the second argues that the op-ed was nothing but problematic.

## Retired Military Officers
One might argue that the logic employed thus far is seemingly plausible for members of the active military, but that the cautionary tone and reservations regarding certain types of political activity should not apply for members of the retired military community. And indeed, there are obvious distinctions between those who are on active duty and those who have retired from the service. These distinctions are made clear in fact in portions of DoD 1344.10 [@department_of_defense_political_2008]. But the fact of the matter is that, especially for retired military officers who achieved general or flag officer rank, these retired officers sometimes carry outsized influence and impact when they engage in activities that approach on the political. During the Presidential Campaign of 2016, Retired Army General Martin Dempsey published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the respective Republican and Democratic conventions. Dempsey wrote, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

@@

> **Yours is the profession of arms, the will to win, the sure knowledge that in war there is no substitute for victory, that if you lose, the Nation will be destroyed, that the very obsession of your public service must be Duty, Honor, Country. Others will debate the controversial issues, national and international, which divide men’s minds. But serene, calm, aloof, you stand as the Nation’s war guardians, as its lifeguards from the raging tides of international conflict, as its gladiators in the arena of battle. For a century and a half you have defended, guarded and protected its hallowed traditions of liberty and freedom, of right and justice. Let civilian voices argue the merits or demerits of our processes of government. Whether our strength is being sapped by deficit financing indulged in too long, by federal paternalism grown too mighty, by power groups grown too arrogant, by politics grown too corrupt, by crime grown too rampant, by morals grown too low, by taxes grown too high, by extremists grown too violent; whether our personal liberties are as firm and complete as they should be. These great national problems are not for your professional participation or military solution. Your guidepost stands out like a tenfold beacon in the night: Duty, Honor, Country** [@macarthur_duty_1962].

\doublespacing
By systematically examining the op-eds written by retired military officers in major US publications over time, it is possible, I contend, to draw inferences about how the concept of a "role belief" - conceptions of one's normative role in politics - has changed among retired US military officers (if at all) [@fitch_armed_1998]. Op-eds are an interesting and appropriate medium to examine if differentiation in "role beliefs" is an object of interest. The first reason why this is the case is because as previously stated, the act of crafting and publishing an op-ed does not necessarily constitute a political activity, though depending on content, they can be. A second and releated reason is because unlike other forms of engaging with the press, such as giving interviews on television, crafting an op-ed requires the author to deliberately prepare and submit a viewpoint for publication. In major newspapers, this is often a competitive process. It is not uncommon to see opinion pieces in prominent newspapers published by heads of state, members of congress, governmental figures, and leaders of industry and academia. A deliberate and carefully-crafted message further underscores the degree to which the author believes the message he or she is sending. It is difficult to "misspeak" during or to be "misquoted" from an opinion-editorial. The author generally agrees with the message he or she is sending.
